---
title: "Research Methods I - Getting Started with R"
author: "Benjamin Graves"
date: "10/14/2021"
output: pdf_document
---

\let\proglang=\textsf
\let\pkg=\emph
\let\code=\texttt

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(psych)
library(psychTools)
library(ggplot2)
## possible data
data("mtcars")
data("sleep")
data("sat.act") ## in psych package
data("epi.bfi") ## psychTools package
```

Hello all. This is an introductory guide to R for Dr. Wood's Research Methods I course. It is designed to get you to try out a few basic but need to know functions in R such as basic statistical analyses and plotting functions. An accompanying video will be uploaded to Dr. Wood's YouTube channel as well and can be found (INSERT LINK HERE). This document assumes you already have \proglang{R} and Rstudio already installed. If not, an instructional video can also be found on Dr. Wood's channel. If you need help with certain functions, or want to know more about them, you can use \code{?FunctionName} to bring up the help file in Rstudio.

# Open a New Script

If this is a new project, it is often a good idea to start up an \proglang{R} script so you can save your code and hardwork to come back to later. To open a new script, you can go up to the tool bar and navigate to 

\begin{center}
\textbf{File > New File > R Script}
\end{center}

Alternatively, below the File drop down menu is a piece of paper with a green plus sign on it. You can click the little arrow beside it to open another drop down menu and select R script from it.


# Set Working Directory

One often overlooked step when starting up an R session is setting up your working directory. This just tells R what folder we want it to look in for files to load in or where to save new files to. It makes it much easier to access files when working on specific projects or to keep everything organized in one folder for a specific class. If you are using Rstudio, you can easily change your working directory by looking up at the task bar and then selecting 

\begin{center}
\textbf{Session > Set Working Directory > Choose Directory}
\end{center}

This will open up a file browser for you to navigate to your folder of choice. Just select it and click okay. You should then see some code in your console window (typically on the left/lower left side of R Studio). You can also use this function, \code{setwd()}, to set the working directory using \proglang{R} functions. 

\ 

```{r wdex, eval=FALSE}
## Code to set the working directory

## NOTE: This file path will be different for everyone and dependent on how
## you have your files set up on your computer

setwd("~/classwork/PSYCH3010")

## If you would like to see what your current WD is
## you can use
getwd()
```

# Installing Packages

We will be using a couple of different packages for the topics below. You might need to install them if this is the first time you are trying to use them. To install packages, we can use the function \code{install.packages()} or by clicking the install button in the \textbf{Packages} tab in the right side of Rstudio.

\ 

```{r install, eval=FALSE}
## Replace 'Package Name' with the name of the package you want to install
install.packages("Package Name", dependencies = TRUE)
```

# Frequency Distributions

The main way to get frequency counts and distributions in \proglang{R} is by using the \code{tables()} an \code{hist()} functions. For this section, we will be using the \code{sat.act} data set found in the \pkg{psych} package.

\ 
```{r tablesfun, comment = NA}
## Load in psych library
library(psych)
## Call up data set
data(sat.act)
## Create factor variables
sat.act$gender <- factor(sat.act$gender, levels = c(1, 2), labels = c('Male', 'Female'))
sat.act$education <- factor(sat.act$education, levels = c(0,1,2,3,4,5),
                            labels = c('< 12 years', 'HS Grad', 'Some College',
                                       'Enrolled College', 'College Grad', 'Grad/Pro Degree'))

## The table function will give us a count of how many
## people are in each group of the gender variable in the data
table(sat.act$gender)
```

```{r histex, comment = NA}
## The breaks argument can be used to increase or decrease the
## width of the bins used in the histogram
hist(sat.act$ACT, breaks = 15, xlab = "ACT Score", main = "Histogram of ACT Scores")
```

# Measures of Central Tendency

The next section covers how to get measures of central tendency and standard deviation for a vector. We will once again be looking at ACT scores using the \code{sat.act} data set.

## Mean

\ 
```{r mean, comment = NA}
mean(sat.act$ACT)
```

## Median

\ 
```{r median, comment = NA}
median(sat.act$ACT)
```

## Mode

\proglang{R} dose not have a base function for calculating the mode of a vector, but there is a package called \pkg{modeest} that will do it for you.

\ 
```{r mode, comment = NA, warning = FALSE, message = FALSE}
library(modeest)

mfv(sat.act$ACT)
```

## Standard Deviation

\ 
```{r sd, comment = NA}
sd(sat.act$ACT)
```

# Welch's Two-Sample t-test

In this section, we conduct Welch's two-sample t-test. This t-test assumes that there are different participants in each group (Independent, Between-subjects) and that the variance of the groups are not equal.

\ 
```{r ttest, comment = NA}
data(sleep)

t.test(extra ~ group, data = sleep)

## NOTE: The defalut argument for t.test is to assume variances are not equal
## If you have tested and see they are, use var.equal = TRUE for Student's t-test

## If data is is wide format, change the arguemtns to be two vectors.
## t.test(sleep$group1, sleep$group2)
```

This function gives us the t-statistic, degrees of freedom, p-value, 95% confidence interval for the difference between the means, and means for each group.

# Cohen's D

We can then get an effect size for the t-test, Cohen's D, using the \pkg{MOTE} package. Since Welch's does not use a pooled variance term, we use the function \code{d.ind.tt()}.

\ 
```{r cohensd, comment = NA}
library(MOTE)

## need t-statistic, group sizes, alpha level
## From sleep data above: t = -1.8608, group 1 n = 10, group 2 n = 10
d.ind.t.t(t = -1.8608, n1 = 10, n2= 10, a = 0.05)
```

This gives us our $d$, lower and upper 95% CI limits, t-test values, and formatting for reporting in APA style.


# Plots

In this section we will be looking at creating various plots.

## Bar Plot

How can we efficiently show the results of the t-test from above? A bar graph showing the means and 95% confidence intervals for each of the two groups would probably be best. We can use the basic plot function for this, but it is pretty involved and difficult process. Instead we will be using \pkg{ggplot2}. 

\ 
```{r bargraph, comment = NA, out.width = '5in', fig.align='center'}
library(ggplot2)

## NOTE: Make sure your grouping variable is a 'factor'
## If not you can turn it into one using factor()

## This code here creates an object to clean up the 
## ugly default ggplot formatting
cleanup <- theme(panel.grid.major = element_blank(), 
                 panel.grid.minor = element_blank(), 
                 panel.background = element_blank(), 
                 axis.line = element_line(colour = "black"))

## Create empty plot object
## In aes, put x-axis variable then y-axis variable
sleepbar <- ggplot(data = sleep, aes(group, extra))

## Use stat summaries to get means and errors bars
sleepbar +
    stat_summary(fun = mean, geom = 'bar', fill = 'white', color = 'black') +
    stat_summary(fun.data = mean_cl_normal, geom = 'errorbar', width = .2) +
    xlab('Group Number') + ylab('Increase in Hours of Sleep') + 
    ggtitle('Sleep Bar Graph') +cleanup
```

The confidence interval here looks a little odd because of the high variance/low sample size of the groups. Generally we would want the error bars to go below the bottom of the bar. In this situation, it might be better to use standard deviation or standard error for the bars. Another example of a bar graph is presented below with the ACT data

\ 
```{r ACTplot, comment = FALSE, out.width = '5in', fig.align='center'}
## Another example using the ACT data
actbar <- ggplot(data = sat.act, aes(education, ACT))

actbar +
    stat_summary(fun = mean, geom = 'bar', fill = 'white', color = 'black') +
    stat_summary(fun.data = mean_cl_normal, geom = 'errorbar', width = .2) +
    xlab('Education Level') + ylab('Average ACT Score') + 
    ggtitle('Average ACT Score Across Education Levels') +cleanup
```

## Box Plot

Box plots are a good way to look at the distribution of a continuous variable across a categorical variable. In this case, we can check to see how the extra hours of sleep are distributed for each of the groups in the sleep data set.

\ 
```{r boxplot, comment = NA, out.width = '5in', fig.align='center'}
sleepbox <- ggplot(sleep, aes(group, extra))

sleepbox +
    geom_boxplot() + xlab('Group') + ylab('Extra Hours of Sleep') +
    ggtitle('Box Plot for Sleep Data') + cleanup
```

## Stem and Leaf Plots

Stem and Leaf plots can be handy to roughly check distributions of data in smaller data sets.

```{r stem, comment = NA}
## The length of the plot can be changed with the 'scale = ' argument
stem(sleep$extra)
```

## Violin Plots

Similar to a box plot is a violin plot. However, instead of just showing quantiles, it plots the density of the distributions for each of the groups.

\ 
```{r violin, comment = NA, out.width = '5in', fig.align='center'}
sleepviolin <- ggplot(sleep, aes(group, extra))

sleepviolin +
    geom_violin() + xlab('Group') + ylab('Extra Hours of Sleep') +
    ggtitle('Box Plot for Sleep Data') + cleanup
```

# Confidence Interval on a Percentage

Confidence intervals for a percentage, as well as a proportion test, can be conducted with the \code{prop.test} function. For this example, lets assume we want to test to see if a coin is fair. Out of $100$ flips the coin lands heads up 62 times.

```{r propCI, comment = NA}
## This function requires the number of successes, 
## total number of observations, and probability to test against
prop.test(x = 62, n = 100, p = .5)
```

# Confidecne interval plots

These plots are handy to compare confidence intervals across several replications of a study. Since we don't have multiple instances of the same study here, we will first simulate 100 data sets and calculate the confidence interval for the mean of each one.

```{r CIplot, comment=NA, fig.height= 8}
## This code creates a loop to generate some data
## and then calculate the confidence interval
lower <- rep(NA, 100)
upper <- rep(NA, 100)
set.seed(587463)
for(i in 1:100) {
  y <- rnorm(100, mean = 5, sd = 5)
  lower[i] <- mean(y) - 1.96 * sd(y) / sqrt(length(y))
  upper[i] <- mean(y) + 1.96 * sd(y) / sqrt(length(y))
}
CIs <- cbind(lower, upper)

## identify intervals not covering mu
## for a 95% CI and 100 samples this should be about 5
ID <- which(!(CIs[1:100, 1] <= 5 & 5 <= CIs[1:100, 2]))

## initialize the plot
plot(0, 
     xlim = c(3, 7), 
     ylim = c(1, 100), 
     ylab = "Sample", 
     xlab = expression(mu), 
     main = "Confidence Intervals")

## set up color vector
colors <- rep(gray(0.6), 100)
colors[ID] <- "red"

## draw reference line at mu=5
abline(v = 5, lty = 2)

## add horizontal bars representing the CIs
for(j in 1:100) {
  lines(c(CIs[j, 1], CIs[j, 2]), 
        c(j, j), 
        col = colors[j], 
        lwd = 2)
}
```

# One-Way ANOVA

One-way ANOVA is handy for comparing more than two groups. For this example, we will be using the \code{sat.act} data set and testing to see if there is a difference between the self-reported ACT score and education level. There are a couple of ways to run the ANOVA, either by using \code{lm()} and \code{anova()} or using \code{ezANOVA()} in the \pkg{ez} package. Both of these require data to be in long format. So, if each column is a group, the data set needs to be reshaped to have a column with all the goups and a column with all the responses.

```{r ANOVA, comment=NA}
## One-way ANOVA using lm() function 
regmodel <- lm(ACT ~ education, data = sat.act)
anova(regmodel)

## Alternatively, use 'ez'
library(ez)

## ez needs a variable with participant numbers
## so create that here
sat.act$partno <- c(1:nrow(sat.act))

## One-way ANOVA
ezANOVA(data = sat.act, 
        dv = ACT, 
        between = education, 
        wid = partno, 
        type = 3, 
        return_aov = TRUE)
```

Using \code{ezANOVA} gives you a little more information, such as Levene's test for homogeneity of variance. Using \code{lm()} and \code{anova()} gives a more traditional ANOVA table output

## *Post-hoc* Tests

Typically, after rejecting the null for the ANOVA, we would do some sort of follow up test to determine where the group differences actually are. This usually consists of doing all possible pairwise group comparisons using a t-test and then correcting for Type I error by either using a stricter cutoff value or correcting the test statistic.

```{r posthoc, comment = NA}
## The argument 'p.adjust.method = ' is what applies the correction
## in this case, we use a Bonferroni corrections on alpha
pairwise.t.test(sat.act$ACT,
                sat.act$education,
                paired = FALSE,
                var.equal = FALSE,
                p.adjust.method = 'bonferroni')
```

# Regresssion with Two Predictors

A regression is run similarly to the first ANOVA option using the \code{lm()} function. all we have to do is specify the formula of the model. Once again, we will be using the \code{sat.act} data set and try to predict ACT using SAT verbal and quantitative scores.

```{r twopredreg, comment = NA}
regmodel <- lm(ACT ~ SATV + SATQ, data = sat.act)
summary(regmodel)
```

## Confidence Intervals for $\beta$

Confidence intervals for the $\beta$ values of the regression model can be obtained using \code{confint()}.

```{r regci, comment=NA}
confint(regmodel, level = 0.95)
```