---
title: "Glossary of R Terminology"
author: "Ben Graves"
date: \today
output: pdf_document
---

\let\proglang=\textsf
\let\pkg=\emph
\let\code=\texttt

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is an evolving list of common terms used when talking about programming in \proglang{R}. I will add new terms as the come up in class, but don't want to overwhelm you all with terms you might not need to know. \proglang{R} code examples are provided below each of the terms. They are designed to be run without any dependencies. For example, each code chunk might call a data set but in a real analysis situation you probably only need to do this once.

## ? or ??

The question mark (\code{?}) can be used to quickly open the help file for a specific function in \proglang{R}. If you want a more general term or phrase, use two question marks (\code{??}) instead.

\  
```{r help, eval=FALSE}
# Provides help for specific function
?mean

## General serch for term or phrase
??correlation 
```


## Argument

Information you provide to a function to make it run. You can see what these are for each function in the help file (use \code{?functionname}). The \code{args()} function will do the same in the console.

\  
```{r arguments, comment=NA}
args(lm)
```

## Assignment Function

Used to assign a name to an object. In \proglang{R}, this is \code{<-}. However, often times you can subsittue this for \code{=} if that is easier. If you want to keep something around for use later, do this. Names can also be whatever you want them to be.

```{r assign, comment = NA}
chickentender <- c(1, 2, 3)
chickentender

cheese = c(4, 5, 6)
cheese
```

## Attributes

Traits of an object such as variable names, object class, row names, and dimension.

```{r attrib, comment = NA}
data(mtcars)
attributes(mtcars)
```

## Character

An element or object consisting of something like words or letters. Must surround the letter or words with quotation makes to denote it as a character.

```{r character, comment = NA}
a <- "Hello!"
a
mode(a)
```

## Class

In short, what the object or element is. For example, a vector, data frame, matrix, or something specific to a function like \code{lm()}.

```{r class, comment = NA}
a <- matrix(c(1, 2, 3, 4), 2, 2)
class(a)

data(mtcars)
class(mtcars)
```

## Data Frame

A data set. Specifically, a set of equal length column vectors that may be of different classes or modes.

```{r dataframe, comment = NA}
data(iris)
head(iris, 3)
```

## Dim (Dimension)

How many rows and columns are in a matrix, data frame, or array.

```{r dims, comment = NA}
data(mtcars)
## The first number is number of rows, the second is number of columns
dim(mtcars)
```

## Element

A specific value in a variable or an item in a vector.

```{r element, comment = NA}
a <- c(4, 5, 6)
a[2]
```

## Factor

A categorical variable with value labels (the "group name") and associated levels (Number assigned to "group").

```{r factors, comment = NA}
Group <- rep(c(1, 2), each = 2)
## Create Factor
Group <- factor(Group, levels = c(1, 2), labels = c("Control", "Treatment"))

## Print Vector
Group

## Labels for levels in factor
levels(Group)
```

## Function

A program that takes arguments and does something with them. Theses are often built in or come in packages, but you can make your own with \code{function()\{\}}

```{r funfun, comment = NA}
MyFunction <- function(x, y, z) {
    result <- x * y + z
    print(result)
}
MyFunction(x = 3, y = 5, z = 2)
```

## Index

Where an element or vector is located in a list, data frame, matrix, or vector. This is based on the dimension of the object. For a two dimensional object, indexing goes by row first then column (ex. \code{data[rows, columns]}).

```{r indexing, comment = NA}
## I want the second element of a vector
a <- c(4, 5, 6)
a[2]

## I want the third variable in a data frame
data(mtcars)
mtcars[ , 3]

## I want the second and fourth row of data
mtcars[c( 2, 4), ]

## What is the MPG for a Dodge Challenger (this works because of row and column names)
mtcars["Dodge Challenger", "mpg"]
```

## Length

The number of elements in a vector or variable. 

```{r longboi, comment = NA}
a <- c(4, 5, 6)
length(a)
```

## Library

This is where \proglang{R} keeps its packages. Also the name of the function to load a package when you need it. You have to do this every time you restart an R session. Must install the package before it can be called or loaded in.

```{r libs, eval = FALSE}
install.packages("psych")
library(psych)
```

## Matrix

A two-dimensional object containing only one type of variable (all numeric or all character).

```{r mat, comment = NA}
m <- matrix(c(1, 2, 3, 4), nrow = 2, ncol = 2, byrow = TRUE)
m
```

## Mode

An object or element's type such as numeric, character, logical, complex.

```{r mode, comment = NA}
a <- 1
mode(a)

b <- "Hello!"
mode(b)

c <- TRUE
mode(c)
```

## NA

A missing value, "Not Available."

```{r missing, comment = NA}
a <- c(1, 2, NA, 4)
is.na(a)
```

## Names

Labels assigned to vectors or elements. Also a function to assign/change variable names.

```{r names, comment = NA}
data(iris)
colnames(iris)
```

## NaN

Another way to specify a missing value. Stands for "Not a Number" and is something undefined mathematically.

```{r nans, comment = NA}
0/0
```

## NULL

The void. The abyss. A vast expanse of nothing. Or an object with zero length and no mode. Assigning null to an object will delete it.

```{r thevoid, comment = NA}
a <- c()
a
```

## Numeric

A vector or element containing only numbers.

```{r numbs, comment = NA}
a <- c(1, 2, 3)
mode(a)
```

## Package

A set of files and functions someone else has already written so you don't have to. Must be installed using \code{install.packages()}.

## Script

A text file containing \proglang{R} code. Ends in ".R" for the file type.

## Subset

To select a series of observations or variables from a larger data set. This can be done through indexing or the \code{subset()} function.

```{r subsetting, comment = NA}
## Select cars with 8 cylinders
data(mtcars)
subset(mtcars, mtcars$cyl == 8)
```

## Vector

Essentially a variable. More specifically a one-dimensional set of values with the same mode.

```{r vects, comment = NA}
a <- c(1, 2, 3, 4, 5)
a
```

## Working Directory (WD)

The file location \proglang{R} is using to look for files to load or where to save files. Best to set this at the start of every \proglang{R} session.

```{r wd, eval = FALSE}
## See what current working directory is
getwd()

## Change your current working directory
setwd("~/filepath")
```
